### 2016-06-28 related-work.tex
Edits.

\paragraph{Reasoning under weak consistency} There have been many
recent proposals focusing on reasoning about programs under weak
consistency~\cite{bailisvldb, alvarocalm, gotsmanpopl16,redblueatc,
redblueosdi, ecinec}. All of them assume a system model that offers a
choice between a \emph{coordination-free} weak consistency level ({\sc
ec} in~\cite{redblueosdi, redblueatc, ecinec, alvarocalm, bailisvldb};
{\sc cc}\footnote{It is worth noting that, although causal consistency
does not require global coordination, it is nonetheless not
\emph{highly available}~\cite{bailishat}.} in~\cite{gotsmanpopl16})
and strong consistency ({\sc sc}), which requires coordination.  They
propose various conditions (\eg \emph{invariant
confluence}~\cite{bailisvldb}, \emph{logical
monotonicity}~\cite{alvarocalm}, \emph{invariant
safety}~\cite{redblueatc}), or proof systems (\eg~rely-guarantee
framework of~\cite{gotsmanpopl16}) to show that an allotment of atomic
and fully isolated operations to these levels preserves application
invariants.  In contrast, we admit weakly isolated transactions (sets
of atomic operations), and our system model accepts
\emph{specifications} of consistency and isolation levels drawn from
an expressive logic. Gotsman et al.~\cite{gotsmanpopl16} adapt
\iso{Parallel Snapshot Isolation} to the aforementioned setting by
interpreting it as a consistency level that serializes writes to
objects. They subsequently propose a dedicated proof rule to prove
program invaraints under this model. By parameterizing our proof
system on isolation specifications, we avoid the need to define a
separate proof rule for each new isolation level.

>>>>

\paragraph{Reasoning under weak consistency} There have been many
recent proposals focusing on reasoning about programs under weak
consistency~\cite{bailisvldb, alvarocalm, gotsmanpopl16,redblueatc,
redblueosdi, ecinec}. All of them assume a system model that offers a
choice between a \emph{coordination-free} weak consistency level ({\sc
ec} in~\cite{redblueosdi, redblueatc, ecinec, alvarocalm, bailisvldb};
{\sc cc} in~\cite{gotsmanpopl16}) and strong consistency ({\sc sc}),
which requires coordination.  They propose various conditions (\eg
\emph{invariant confluence}~\cite{bailisvldb}, \emph{logical
monotonicity}~\cite{alvarocalm}, \emph{invariant
safety}~\cite{redblueatc}), or proof systems (\eg~rely-guarantee
framework of~\cite{gotsmanpopl16}) to show that an allotment of atomic
and fully isolated operations to these levels preserves application
invariants.  In contrast, we admit weakly isolated transactions (sets
of atomic operations), and our system model accepts
\emph{specifications} of consistency and isolation levels drawn from
an expressive logic. Gotsman et al.~\cite{gotsmanpopl16} adapt
\iso{Parallel Snapshot Isolation} to the aforementioned setting by
interpreting it as a consistency level that serializes writes to
objects. They subsequently propose a dedicated proof rule to prove
program invaraints under this model. By parameterizing our proof
system on isolation specifications, we avoid the need to define a
separate proof rule for each new isolation level.

### 2016-06-28 store-consistency.tex
Related work done.
Edits.

In practice, relational databases resolve such conflicts by
prioritizing weak isolation (thus, concurrency and performance) over
strong consistency, so the execution traces do not necessarily satisfy
{\sc sc}. In particular, visibility constraints imposed by {\sc sc}
are violated iff they are found to be in conflict with the constraints
imposed by a transaction isolation level. In the context of
the aforementioned example, $\eta_1$ is not made visible to $\eta_2$
because doing so would violate $\I_c$. However, if $\I_c$ is true,
then the store makes $\eta_1$ visible to $\eta_2$ to honor its
consistency commitment\footnote{The term \emph{recency
    commitment}~\cite{bailisvldb} is often used in practice
  to capture the best-effort nature of {\sc sc}.}. We generalize this
approach to any $\I_s$ and $\I_c$ by defining a \emph{maximum
  visiblity principle} to determine an acceptable weakening of $\I_s$
in case of a conflict with $\I_c$.  The principle requires the
weakened consistency guarantee ($\I_s'$) of the store to enforce all
visibility relationships imposed by the actual consistency guarantee
($\I_s$), unless enforcing such a relationship violates $\I_c$.
% The formal definition of the principle is relegated
% to the supplementary in the interest of space, but it can be
% understood in the context of an {\sc sc} store, where it weakens the
% store invariant to the following:
Formally:
\begin{definition}
$\I_s' : \E \rightarrow \Prop$ is said to be a maximum visibility
weakening of $\I_s : \E \rightarrow \Prop$ if and only if:
\begin{itemize}
  \item $\I_s'$ is weaker than $\I_s$: 
      $\forall\E.~ \I_s(\E) \Rightarrow \I_s'(\E)$, and
  \item In every trace $\E$ that satisfies $\I_c$, and for every pair
  of effects $\eta_1$ and $\eta_2$ in $\E$, if $\I_s(\E)$ requires
  $\eta_1$ to be visible to $\eta_2$, then so does $\I_s'(\E)$ unless
  extending $\E$ with $\visZ(\eta_1,\eta_2)$ violates
  $\I_c$\footnote{\GK{ToDo: consider other well-formedness conditions
  on trace, such as acyclicity of $\visZ$ and $\soZ$. Are they needed
  (considering that the machine never violates them)? Encode the
  specifications in Z3 and make sure they are consistent.}}:
  \begin{smathpar}
  \begin{array}{l}
  \forall\E,\eta_1,\eta_2.~ \I_c(\E) \Rightarrow (\I_s(\E)
    \Rightarrow \underE{\eta_1 \visar \eta_2}) \Rightarrow \\
    \hspace*{0.5in}(\I_s'(\E) \Rightarrow \underE{\eta_1 \visar
    \eta_2} \disj \neg\I_c(\E\,\cup\,(\emptyset,\{(\eta_1,\eta_2)\})))
  \end{array}
  \end{smathpar}
\end{itemize}
\end{definition}
Applying this principle, we can weaken {\sc sc} to obtain the
following store trace invariant ($\I_s$) for an {\sc sc} store whose
isolation constraints are captured by $\I_c$:
\begin{smathpar}
\begin{array}{lcl}
\I_s(\E) & = & \forall \eta_1,\eta_2.\, \{\eta_1,\eta_2\},
    \subseteq \E.\A \conj \id(\eta_1) <
    \id(\eta_2) \\
    & & \hspace*{0.5in} \Rightarrow 
      \underE{\eta_1 \visar \eta_2} \disj \neg\I_c(\E
    \cup (\emptyset,\{(\eta_1,\eta_2)\}))\\
\end{array}
\end{smathpar}
$\I_s$ requires a trace $\E$ to satisfy the visibility constraints of
{\sc sc} except in cases where they are in conflict with $\I_c$.
Instantiating the parameter $\I$  with $\I_s \wedge \I_c$ in
Fig.~\ref{fig:txnimp} results in an operational semantics that
describes many kinds of relational database-like stores.

@@

\subsection{Weakly Consistent Replication}

@@

Reasoning about weakly-consistent replication involves reasoning in
terms of \emph{replicas} of data. The primary challenge in this
setting is to ensure that the assumptions made and guarantees enforced
by an operation at one replica carry over to other replicas that merge
their effects, thus preserving the overall integrity of the system.
In~\cite{gotsmanpopl16}, this challenge is partly addressed by
imposing restrictions on how various replica states differ, i.e., by
fixing a system model with a stronger baseline consistency ({\sc cc})
than {\sc ec}. This unfortunately restricts the reasoning approach
from being applied to stores that provide guarantees weaker than
causal consistency (e.g., \cite{bayou,pldi15}).

Notably, our view of replication however does not explicitly involve
replicas.  Fig.~\ref{fig:ec-theirs-vs-ours} contrasts our model of
weakly-consistent replication with a conventional replica-based
model.  Under our model, the notion of a replica is subsumed by the
concept of visibility; a replica is defined by the subset ($S$) of
global state ($\E.\A$) that an operation witnesses. Constraints over
replica states therefore manifest as constraints over a specific visibility
relation.  For example, instead of requiring the store to be causally
consistent, an operation can request to witness a causally consistent
subset of the state; such demands can be made via the trace invariant
$\I$. For a precondition ($P$) of the operation to be useful, it has
to be an assertion over every causally consistent subset of the global
state.  Since any replica that eventually executes the operation has
to expose one such subset ($S$), the precondition is guaranteed to
hold regardless of the replica. There is however one problem with this
explanation - by considering subsets of just one global state, it
ignores the fact that the global state (hence, the replica states)
change during the execution of the operation. To account for this
change,~\cite{gotsmanpopl16} distinguishes between an \emph{effect
  generation} event at one replica and an \emph{effect merge} event at
another replica, and requires that certain invariants be preserved
between these two events at different replicas. Our framework folds
all such machinery into a stability condition (\S\ref{sec:rely-guarantee}).
Since any change to the global state during the execution of the
operation is an interference, and $P$ is required to be stable with
respect to any such interference, it follows that $P$ is valid on
every replica, thus ensuring that assumptions made at the generation
event are also valid at the merge event.

>>>>

In practice, relational databases resolve such conflicts by
prioritizing weak isolation (thus, concurrency and performance) over
strong consistency, so the execution traces do not necessarily satisfy
{\sc sc}. In particular, visibility constraints imposed by {\sc sc}
are violated iff they are found to be in conflict with the constraints
imposed by a transaction isolation level. In the context of
the aforementioned example, $\eta_1$ is not made visible to $\eta_2$
because doing so would violate $\I_c$. However, if $\I_c$ is true,
then the store makes $\eta_1$ visible to $\eta_2$ to honor its
consistency commitment\footnote{The term \emph{recency
    commitment}~\cite{bailishat} is often used in practice
  to capture the best-effort nature of {\sc sc}.}. We generalize this
approach to any $\I_s$ and $\I_c$ by defining a \emph{maximum
  visiblity principle} to determine an acceptable weakening of $\I_s$
in case of a conflict with $\I_c$.  The principle requires the
weakened consistency guarantee ($\I_s'$) of the store to enforce all
visibility relationships imposed by the actual consistency guarantee
($\I_s$), unless enforcing such a relationship violates $\I_c$.
% The formal definition of the principle is relegated
% to the supplementary in the interest of space, but it can be
% understood in the context of an {\sc sc} store, where it weakens the
% store invariant to the following:
Formally:
\begin{definition}
$\I_s' : \E \rightarrow \Prop$ is said to be a maximum visibility
weakening of $\I_s : \E \rightarrow \Prop$ if and only if:
\begin{itemize}
  \item $\I_s'$ is weaker than $\I_s$: 
      $\forall\E.~ \I_s(\E) \Rightarrow \I_s'(\E)$, and
  \item In every trace $\E$ that satisfies $\I_c$, and for every pair
  of effects $\eta_1$ and $\eta_2$ in $\E$, if $\I_s(\E)$ requires
  $\eta_1$ to be visible to $\eta_2$, then so does $\I_s'(\E)$ unless
  extending $\E$ with $\visZ(\eta_1,\eta_2)$ violates
  $\I_c$\footnote{\GK{ToDo: consider other well-formedness conditions
  on trace, such as acyclicity of $\visZ$ and $\soZ$. Are they needed
  (considering that the machine never violates them)? Encode the
  specifications in Z3 and make sure they are consistent.}}:
  \begin{smathpar}
  \begin{array}{l}
  \forall\E,\eta_1,\eta_2.~ \I_c(\E) \Rightarrow (\I_s(\E)
    \Rightarrow \underE{\eta_1 \visar \eta_2}) \Rightarrow \\
    \hspace*{0.5in}(\I_s'(\E) \Rightarrow \underE{\eta_1 \visar
    \eta_2} \disj \neg\I_c(\E\,\cup\,(\emptyset,\{(\eta_1,\eta_2)\})))
  \end{array}
  \end{smathpar}
\end{itemize}
\end{definition}
Applying this principle, we can weaken {\sc sc} to obtain the
following store trace invariant ($\I_s$) for an {\sc sc} store whose
isolation constraints are captured by $\I_c$:
\begin{smathpar}
\begin{array}{lcl}
\I_s(\E) & = & \forall \eta_1,\eta_2.\, \{\eta_1,\eta_2\},
    \subseteq \E.\A \conj \id(\eta_1) <
    \id(\eta_2) \\
    & & \hspace*{0.5in} \Rightarrow 
      \underE{\eta_1 \visar \eta_2} \disj \neg\I_c(\E
    \cup (\emptyset,\{(\eta_1,\eta_2)\}))\\
\end{array}
\end{smathpar}
$\I_s$ requires a trace $\E$ to satisfy the visibility constraints of
{\sc sc} except in cases where they are in conflict with $\I_c$.
Instantiating the parameter $\I$  with $\I_s \wedge \I_c$ in
Fig.~\ref{fig:txnimp} results in an operational semantics that
describes many kinds of relational database-like stores.

@@

\subsection{Weakly Consistent Replication}
\label{sec:replication}

@@

Reasoning about weakly-consistent replication involves reasoning in
terms of \emph{replicas} of data. The primary challenge in this
setting is to ensure that the assumptions made and guarantees enforced
by an operation at one replica carry over to other replicas that merge
their effects, thus preserving the overall integrity of the system.
In~\cite{gotsmanpopl16}, this challenge is partly addressed by
imposing restrictions on how various replica states differ, i.e., by
fixing a system model with a stronger baseline consistency ({\sc cc})
than {\sc ec}. This unfortunately restricts the reasoning approach
from being applied to stores (e.g., \cite{bayou,pldi15}) that provide
guarantees weaker than causal consistency, such as causal visibility
or read-my-writes~\cite{zoo}. Causal consistency is not baseline
consistency in these stores because it is not \emph{highly
available}~\cite{bailishat}.

Notably, our view of replication however does not explicitly involve
replicas.  Fig.~\ref{fig:ec-theirs-vs-ours} contrasts our model of
weakly-consistent replication with a conventional replica-based
model.  Under our model, the notion of a replica is subsumed by the
concept of visibility; a replica is defined by the subset ($S$) of
global state ($\E.\A$) that an operation witnesses. Constraints over
replica states therefore manifest as constraints over a specific visibility
relation.  For example, instead of requiring the store to be causally
consistent, an operation can request to witness a causally consistent
subset of the state; such demands can be made via the trace invariant
$\I$. For a precondition ($P$) of the operation to be useful, it has
to be an assertion over every causally consistent subset of the global
state.  Since any replica that eventually executes the operation has
to expose one such subset ($S$), the precondition is guaranteed to
hold regardless of the replica. There is however one problem with this
explanation - by considering subsets of just one global state, it
ignores the fact that the global state (hence, the replica states)
change during the execution of the operation. To account for this
change,~\cite{gotsmanpopl16} distinguishes between effect generation
event at one replica $r_1$ and effect merge event at replica $r_2$,
and requires that only \emph{non-conflicting} operations execute
between these two events at $r_2$, and that they preserve certain
invariants. Our framework folds all such machinery into a stability
condition predicated on $\I$ (\S\ref{sec:rely-guarantee}).  Since any
change to the global state during the execution of the operation is an
interference, and $P$ is required to be stable with respect to any
such interference, it follows that $P$ is valid on every replica, thus
ensuring that assumptions made at the generation event are also valid
at the merge event.
